---
title: OSH demos
author: "AS Huffmyer"
date: '2026'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays conducted at VIMS for OSH demos

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
library(pracma)
library(rptR)
library(fda)
library(nls2)
library(purrr)
library(broom)
library(pheatmap)
library(vegan)
library(ggridges)
library(ggpubr)
library(broom.mixed)
```

# Resazurin Trials   

## Load data 

```{r}
# Set the folder path
folder_path <- "data/OSH-demo/plate-files/"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.xlsx$", full.names = TRUE, recursive=TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No xlsx files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read Excel file, skipping the first 9 rows
  data <- read_excel(file, skip = 10, col_names = FALSE)

  # Keep only the first 8 rows (row 1–8 after - skip)
  data <- data[1:8, ]
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:12))
  
  #remove first column
  #data<-data%>%select(!"Row")
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      plate = str_extract(file_name, "Plate\\d+"),  # Extract "plateX"
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, date, plate, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/OSH-demo/metadata.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)

full_data<-left_join(combined_data, metadata, by=c("date", "well", "plate"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_xlsx("data/OSH-demo/size.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data. 

```{r}
str(full_data)
str(size)

full_data<-left_join(full_data, size, by=c("date", "plate", "well"))
head(full_data)

full_data<-full_data%>%
  select(!Source)%>%
  select(!FileName)

head(full_data)
```

## Prep the data 

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=treatment, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate size normalized fluorescence at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, plate, well, treatment)%>%
  arrange(date, plate, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=treatment, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate mean change in blank at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, timepoint)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank))+
  facet_wrap(~date)+
  geom_point()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(fluorescence.corr=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=treatment, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=length.mm, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  scale_colour_gradient(low="darkblue", high="red")+
  geom_point()+
  geom_line()+
  theme_classic()
```

Size normalize data.
```{r}
#normalize by length
full_data<-full_data%>%
  mutate(fluorescence.corr.mm=fluorescence.corr/length.mm)
```

Plot again. 

Plot by area normalized. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm, colour=treatment, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Plot by area normalized colored by size. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm, colour=length.mm, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  scale_colour_gradient(low="darkblue", high="red")+
  geom_point()+
  geom_line()+
  theme_classic()
```

Remove unnecessary columns. 

```{r}

full_data<-full_data%>%
  #select(!resazurin_counts)%>%
  select(!type)#%>%
  #select(!mean_blank)
  #select(!length.mm)%>%
  #select(!fluorescence.corr)%>%
  #select(!fluorescence.norm)

full_data<-full_data%>%
  rename(value=fluorescence.corr.mm)
```

### Remove oversaturated samples

Remove any sample that reached > 3000 raw value at any point during the trials, these were over saturated and data is unreliable. 

```{r}
hist(full_data$resazurin_counts)

full_data%>%
  filter(resazurin_counts>3000)%>%
  group_by(date, plate, well)%>%
  mutate(unique=paste(date, plate, well))

#there are no samples that went above the threshold 

list<-full_data%>%
  filter(resazurin_counts>3000)%>%
  group_by(date, plate, well)%>%
  mutate(unique=paste(date, plate, well))%>%
  pull(unique)%>%
  unique()
```

```{r}
full_data<-full_data%>%
  mutate(unique=paste(date, plate, well))%>%
  filter(!unique %in% list)
```

View size range. 
```{r}
hist(full_data$length.mm)
```

Plot again. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=treatment, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

## Models and Plots

Plot data. 
```{r}
plot1<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=treatment, group=interaction(date, treatment, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic();plot1
```

Run a model on metabolic rates. 

```{r}
model<-full_data%>%
  lmer(sqrt(value) ~ treatment * timepoint + (1|well:plate) + (1|plate), data=.)

anova(model)
qqPlot(residuals(model))
leveneTest(sqrt(value) ~ treatment * timepoint, data=full_data)
```
Significant treatment x time effect - different metabolic rates. 

Plot raw data for each ploidy 

```{r}
plot2<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=treatment, group=interaction(date, treatment)))+
  geom_smooth(method="lm")+
  theme_classic();plot2
```

View sample size. 
```{r}
full_data%>%
  group_by(treatment, timepoint)%>%
  summarise(length(value))
```

Plot model estimates for metabolic rates for each ploidy 

```{r}
#predict data 
pred_data <- augment(model, full_data) %>%
  mutate(predicted_value_ploidy = (.fitted))  # back-transform to original scale

#plot with back transformed data 
plot2e<-ggplot(pred_data, aes(x = timepoint, y = predicted_value_ploidy^2, color = treatment, group = treatment, fill = treatment)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  geom_vline(xintercept = 3, linetype="dashed")+
  stat_summary(
    fun.data = mean_se,
    geom = "ribbon",
    alpha = 0.2,
    color = NA
  ) +
  labs(
    title = "Modeled Metabolic Rate",
    x = "Timepoint",
    y = "Fold Change in Fluorescence"
  ) +
  theme_classic() +
  guides(fill = "none");plot2e  
```

## Area under the curve 

Area under the curve (AUC): Summarizes total metabolic activity per oyster.

```{r}
aucs <- full_data %>%
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, plate, well, treatment) %>%
  summarise(AUC = trapz(time, value))
```

```{r}
plot10<-ggplot(aucs, aes(x = treatment, y = AUC, colour = treatment)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(title = "AUC by Treatment", x = "Treatment", y = "AUC") +
  theme_classic();plot10

plot10a<-aucs%>%
  group_by(treatment)%>%
  summarise(mean=mean(AUC, na.rm=TRUE), se=(sd(AUC, na.rm=TRUE))/sqrt(length(AUC)))%>%
  
  ggplot(aes(x = reorder(treatment, mean), y = mean, colour = treatment)) +
  geom_point()+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=0.1)+
  labs(title = "Total metabolic activity (AUC)", x = "Treatment", y = "AUC") +
  theme_classic();plot10a
```

Model the data. 

```{r}
hist(sqrt(aucs$AUC))

model<-lmer(sqrt(AUC) ~ treatment + (1|plate), data=aucs)

anova(model)
qqPlot(residuals(model))
```

## Ridge plots 

View distrubution of fluorescence values over time. 

```{r}
plot14<-ggplot(full_data, aes(x = value, y = as.factor(timepoint), fill = treatment)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Distribution of Fluorescence by Treatment")+
  theme_classic();plot14
```
