---
title: "resazurin-curve-features-all-experiments"
author: "Sam White, Ariana Huffmyer"
date: "2026-07-04"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options:
  chunk_output_type: console
---

# Background

This notebook applies the curve-feature comparison approach across all
resazurin experiment-level analyses. It uses the existing
`metabolism.csv` outputs.

## Expected inputs

| Path | Description |
|------|-------------|
| `output/metabolism.csv` | Per-well per-timepoint blank-corrected and size-normalized metabolism data |

## Expected outputs

All outputs are written to
`output/resazurin-curves/`.

| File | Description |
|------|-------------|
| `curve_features.csv` | Per-individual curve traits for every included experiment |
| `curve_feature_family_summary.csv` | Family-level summaries for each curve trait |
| `curve_feature_family_stats.csv` | Per-experiment feature ranking by family separation |
| `curve_feature_pairwise_stats.csv` | Tukey-adjusted family comparisons for top-ranked curve traits |
| `curve_feature_trajectory_classes.csv` | Trajectory-class counts and proportions by experiment and family |
| `included_experiments.csv` | Experiment output folders included in the analysis |
| `figures/` | Cross-experiment ranking, heatmap, best-feature, and trajectory plots |

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  results = "hold"
)
```

```{r libraries}
library(tidyverse)
library(colorspace)
library(here)
```

# Helper Functions

```{r helper-fns}
safe_divide <- function(num, den) {
  if_else(is.finite(num) & is.finite(den) & abs(den) > .Machine$double.eps,
          num / den, NA_real_)
}

trapezoid_auc <- function(time_hr, value) {
  ok <- is.finite(time_hr) & is.finite(value)
  t <- time_hr[ok]
  v <- value[ok]
  if (length(t) < 2) return(NA_real_)
  ord <- order(t)
  t <- t[ord]
  v <- v[ord]
  sum(diff(t) * (head(v, -1) + tail(v, -1)) / 2)
}

segment_auc <- function(time_hr, value) {
  ok <- is.finite(time_hr) & is.finite(value)
  t <- time_hr[ok]
  v <- value[ok]
  if (length(t) < 2) {
    return(tibble(t_start = numeric(), t_end = numeric(),
                  t_mid = numeric(), auc = numeric()))
  }
  ord <- order(t)
  t <- t[ord]
  v <- v[ord]
  tibble(
    t_start = head(t, -1),
    t_end   = tail(t, -1),
    t_mid   = (head(t, -1) + tail(t, -1)) / 2,
    auc     = diff(t) * (head(v, -1) + tail(v, -1)) / 2
  )
}

classify_trajectory <- function(slopes, slope_tol = 1e-12) {
  if (length(slopes) == 0 || all(!is.finite(slopes))) return(NA_character_)
  s <- slopes[is.finite(slopes)]
  signs <- case_when(
    s >  slope_tol ~  1L,
    s < -slope_tol ~ -1L,
    TRUE           ~  0L
  )
  nonzero <- signs[signs != 0L]
  if (length(nonzero) == 0) return("stable")
  if (all(nonzero > 0)) return("monotonic_increase")
  if (all(nonzero < 0)) return("monotonic_decrease")
  if (any(nonzero > 0) && any(nonzero < 0)) {
    first_neg <- which(nonzero < 0)[1]
    first_pos <- which(nonzero > 0)[1]
    if (first_pos < first_neg) return("rise_then_depress")
    if (first_neg < first_pos) return("depress_then_recover")
  }
  "complex"
}

extract_curve_features_one <- function(df, value_col, metric_label) {
  df <- df %>%
    filter(is.finite(time_hr), is.finite(.data[[value_col]])) %>%
    arrange(time_hr)

  if (nrow(df) < 2) return(tibble())

  t <- df$time_hr
  v <- df[[value_col]]
  dt <- diff(t)
  dv <- diff(v)
  valid_slope <- is.finite(dt) & dt > 0 & is.finite(dv)
  slopes <- dv[valid_slope] / dt[valid_slope]
  slope_midpoints <- (head(t, -1) + tail(t, -1)) / 2
  slope_midpoints <- slope_midpoints[valid_slope]
  seg_auc <- segment_auc(t, v)
  split_time <- median(seg_auc$t_mid, na.rm = TRUE)
  early_auc <- seg_auc %>%
    filter(t_mid <= split_time) %>%
    summarise(x = sum(auc, na.rm = TRUE), .groups = "drop") %>%
    pull(x)
  late_auc <- seg_auc %>%
    filter(t_mid > split_time) %>%
    summarise(x = sum(auc, na.rm = TRUE), .groups = "drop") %>%
    pull(x)

  if (length(early_auc) == 0) early_auc <- NA_real_
  if (length(late_auc) == 0) late_auc <- NA_real_

  baseline_value <- v[1]
  final_value <- v[length(v)]
  peak_idx <- which.max(v)
  trough_idx <- which.min(v)
  peak_value <- v[peak_idx]
  trough_value <- v[trough_idx]
  peak_after_baseline <- peak_value - baseline_value
  depression_abs <- if (peak_idx < length(v)) {
    max(peak_value - final_value, 0, na.rm = TRUE)
  } else {
    0
  }

  vmax_idx <- if (length(slopes) > 0) which.max(slopes) else NA_integer_
  min_slope_idx <- if (length(slopes) > 0) which.min(slopes) else NA_integer_
  curvature <- diff(slopes)
  inflection_idx <- if (length(curvature) > 0) {
    which.max(abs(curvature)) + 1L
  } else {
    NA_integer_
  }

  tibble(
    value_metric = metric_label,
    baseline_value = baseline_value,
    final_value = final_value,
    peak_value = peak_value,
    trough_value = trough_value,
    time_to_peak = t[peak_idx],
    time_to_trough = t[trough_idx],
    auc_total = trapezoid_auc(t, v),
    auc_early = early_auc,
    auc_late = late_auc,
    delta_auc_late_minus_early = late_auc - early_auc,
    late_early_auc_ratio = safe_divide(late_auc, early_auc),
    initial_slope = slopes[1],
    vmax = if (length(slopes) > 0) max(slopes, na.rm = TRUE) else NA_real_,
    time_to_vmax = if (!is.na(vmax_idx)) slope_midpoints[vmax_idx] else NA_real_,
    min_slope = if (length(slopes) > 0) min(slopes, na.rm = TRUE) else NA_real_,
    time_to_min_slope = if (!is.na(min_slope_idx)) {
      slope_midpoints[min_slope_idx]
    } else {
      NA_real_
    },
    inflection_time = if (!is.na(inflection_idx)) {
      slope_midpoints[inflection_idx]
    } else {
      NA_real_
    },
    metabolic_scope = peak_after_baseline,
    final_delta = final_value - baseline_value,
    depression_abs = depression_abs,
    metabolic_depression_index = safe_divide(depression_abs,
                                             abs(peak_after_baseline)),
    resilience_ratio = safe_divide(final_value, peak_value),
    stability_cv = safe_divide(sd(v, na.rm = TRUE), abs(mean(v, na.rm = TRUE))),
    trajectory_class = classify_trajectory(slopes),
    n_timepoints = nrow(df)
  )
}

compute_curve_features <- function(df, value_col, metric_label) {
  id_vars <- intersect(
    c("trace_id", "family_id_group", "phenotype",
      "round_group", "cup_id_group", "sample_id_group"),
    names(df)
  )

  df %>%
    filter(!is.na(family_id_group), is.finite(.data[[value_col]])) %>%
    group_by(across(all_of(id_vars))) %>%
    group_modify(~ extract_curve_features_one(.x, value_col, metric_label)) %>%
    ungroup()
}

tidy_tukey <- function(model, term) {
  tk <- tryCatch(TukeyHSD(aov(model), which = term)[[term]],
                 error = function(e) NULL)
  if (is.null(tk)) return(tibble())
  as_tibble(tk, rownames = "contrast") %>%
    rename(
      estimate = diff,
      p.value = `p adj`
    ) %>%
    mutate(SE = NA_real_, df = NA_real_, t.ratio = NA_real_) %>%
    select(contrast, estimate, SE, df, t.ratio, p.value)
}

run_feature_family_stats <- function(df) {
  df <- df %>%
    filter(is.finite(feature_value), !is.na(family_id_group)) %>%
    mutate(family = factor(family_id_group))

  if (nrow(df) < 3 || n_distinct(df$family) < 2) return(tibble())

  model <- lm(feature_value ~ family, data = df)
  aov_tab <- anova(model)

  ss_family <- aov_tab["family", "Sum Sq"]
  df_family <- aov_tab["family", "Df"]
  ms_error <- aov_tab["Residuals", "Mean Sq"]
  ss_total <- sum(aov_tab$`Sum Sq`, na.rm = TRUE)
  eta_sq <- ss_family / ss_total
  omega_sq <- (ss_family - df_family * ms_error) / (ss_total + ms_error)

  fam_means <- df %>%
    group_by(family) %>%
    summarise(mean = mean(feature_value), .groups = "drop")

  pair_grid <- tidyr::crossing(fam1 = fam_means$family,
                               fam2 = fam_means$family) %>%
    filter(as.character(fam1) < as.character(fam2)) %>%
    left_join(fam_means, by = c("fam1" = "family")) %>%
    rename(mean1 = mean) %>%
    left_join(fam_means, by = c("fam2" = "family")) %>%
    rename(mean2 = mean) %>%
    mutate(abs_d = abs(mean1 - mean2) / sqrt(ms_error))

  kw <- kruskal.test(feature_value ~ family, data = df)

  tibble(
    n = nrow(df),
    n_family = n_distinct(df$family),
    f_statistic = aov_tab["family", "F value"],
    p_value = aov_tab["family", "Pr(>F)"],
    kruskal_p_value = kw$p.value,
    eta_sq = eta_sq,
    omega_sq = max(omega_sq, 0, na.rm = TRUE),
    max_abs_pairwise_d = max(pair_grid$abs_d, na.rm = TRUE)
  )
}

run_feature_pairwise <- function(df) {
  df <- df %>%
    filter(is.finite(feature_value), !is.na(family_id_group)) %>%
    mutate(family = factor(family_id_group))

  if (nrow(df) < 3 || n_distinct(df$family) < 2) return(tibble())

  model <- lm(feature_value ~ family, data = df)
  tidy_tukey(model, "family")
}

make_palette <- function(n) {
  okabe_ito_7 <- c(
    "#E69F00", "#56B4E9", "#009E73", "#F0E442",
    "#0072B2", "#D55E00", "#CC79A7"
  )
  if (n == 0L) return(character(0))
  if (n <= length(okabe_ito_7)) return(okabe_ito_7[seq_len(n)])
  colorspace::qualitative_hcl(n, palette = "Dynamic")
}
```

# Load Experiment Outputs

```{r discover-inputs}
out_dir <- here("output", "resazurin-curves")
fig_dir <- file.path(out_dir, "figures")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

metabolism_path <- here("output", "metabolism.csv")
stopifnot(file.exists(metabolism_path))

included_experiments <- tibble(
  metabolism_path = metabolism_path,
  experiment      = "vims-resazurin"
)

included_experiments
```

```{r load-metabolism}
metabolism_all <- read_csv(
  metabolism_path,
  col_types = cols(.default = "c"),
  show_col_types = FALSE
) %>%
  rename(
    time_hr         = timepoint,
    family_id_group = family,
    corrected_fc    = value,
    trace_id        = unique
  ) %>%
  mutate(
    family_id_group = as.character(family_id_group),
    across(c("corrected_fc", "time_hr"), ~ suppressWarnings(as.numeric(.x))),
    trace_id        = as.character(trace_id)
  )

metabolism_all <- metabolism_all %>%
  mutate(
    family_id_group = str_to_lower(trimws(family_id_group)),
    phenotype       = str_to_lower(trimws(phenotype))
  )

experiment_overview <- metabolism_all %>%
  group_by(phenotype) %>%
  summarise(
    n_rows        = n(),
    n_individuals = n_distinct(trace_id),
    n_families    = n_distinct(family_id_group, na.rm = TRUE),
    min_time_hr   = min(time_hr, na.rm = TRUE),
    max_time_hr   = max(time_hr, na.rm = TRUE),
    n_timepoints  = n_distinct(time_hr),
    .groups = "drop"
  )

print(experiment_overview)
```

# Curve Feature Extraction

```{r curve-features}
value_metric_cols <- c(
  corrected_fc = "corrected_fc",
  setNames(
    names(metabolism_all)[str_detect(names(metabolism_all),
                                     "^metabolism_per_")],
    names(metabolism_all)[str_detect(names(metabolism_all),
                                     "^metabolism_per_")]
  )
)
value_metric_cols <- value_metric_cols[value_metric_cols %in% names(metabolism_all)]

curve_features <- imap_dfr(
  value_metric_cols,
  ~ compute_curve_features(metabolism_all, .x, .y)
)

numeric_feature_cols <- curve_features %>%
  select(where(is.numeric)) %>%
  select(-any_of("n_timepoints")) %>%
  names()

curve_feature_long <- curve_features %>%
  pivot_longer(
    cols = all_of(numeric_feature_cols),
    names_to = "feature",
    values_to = "feature_value"
  ) %>%
  filter(is.finite(feature_value))

trajectory_features <- curve_features %>%
  count(phenotype, value_metric, family_id_group, trajectory_class, name = "n") %>%
  group_by(phenotype, value_metric, family_id_group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

str(curve_features)
```

# Family Summaries

```{r family-summary}
curve_feature_family_summary <- curve_feature_long %>%
  group_by(phenotype, value_metric, feature, family_id_group) %>%
  summarise(
    n = n(),
    mean = mean(feature_value, na.rm = TRUE),
    sd = sd(feature_value, na.rm = TRUE),
    se = sd / sqrt(n),
    median = median(feature_value, na.rm = TRUE),
    iqr = IQR(feature_value, na.rm = TRUE),
    .groups = "drop"
  )

print(curve_feature_family_summary)
```

# Rank Family-Discriminating Features

For each experiment and numeric feature, the model
`feature_value ~ family_id_group` is fit with individual oysters as
observational units. Features are ranked by a combined score favoring stronger
family effect size (`eta_sq`) and lower BH-adjusted p-value. This is intended
as a screening/ranking analysis rather than a single confirmatory model.
`baseline_value` is retained in the full feature table but excluded from the
ranking because the goal here is to compare curve shape, capacity,
responsiveness, and depression-like behavior rather than the normalized
starting reference.

```{r feature-stats}
curve_feature_family_stats <- curve_feature_long %>%
  filter(feature != "baseline_value") %>%
  group_by(phenotype, value_metric, feature) %>%
  group_modify(~ run_feature_family_stats(.x)) %>%
  ungroup() %>%
  group_by(phenotype, value_metric) %>%
  mutate(
    p_adj_bh = p.adjust(p_value, method = "BH"),
    rank_within_metric = min_rank(p_adj_bh),
    rank_score = dense_rank(desc(eta_sq)) + dense_rank(p_adj_bh)
  ) %>%
  ungroup() %>%
  arrange(phenotype, p_adj_bh, desc(eta_sq), desc(max_abs_pairwise_d))

best_curve_features <- curve_feature_family_stats %>%
  group_by(phenotype, value_metric) %>%
  slice_min(order_by = rank_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(phenotype, value_metric)

top_curve_features <- curve_feature_family_stats %>%
  group_by(phenotype, value_metric) %>%
  slice_min(order_by = rank_score, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(phenotype, value_metric, rank_score)

print(best_curve_features)
```

```{r pairwise-stats}
curve_feature_pairwise_stats <- curve_feature_long %>%
  semi_join(top_curve_features, by = c("phenotype", "value_metric", "feature")) %>%
  group_by(phenotype, value_metric, feature) %>%
  group_modify(~ run_feature_pairwise(.x)) %>%
  ungroup()

print(curve_feature_pairwise_stats)
```

# Most Distinguishing Aspects

```{r best-feature-print, results='asis'}
cat("## Best curve feature per experiment and metric\n\n")
cat(knitr::kable(best_curve_features, digits = 5, format = "pipe"), sep = "\n")
cat("\n\n## Top five curve features per experiment and metric\n\n")
cat(knitr::kable(top_curve_features, digits = 5, format = "pipe"), sep = "\n")
```

```{r feature-rank-plot, fig.width=12, fig.height=10}
rank_plot_df <- curve_feature_family_stats %>%
  group_by(phenotype, value_metric) %>%
  slice_max(order_by = eta_sq, n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(feature = fct_reorder(feature, eta_sq))

p_feature_rank <- ggplot(rank_plot_df,
    aes(x = eta_sq, y = feature, fill = p_adj_bh < 0.05)) +
  geom_col(width = 0.75) +
  facet_grid(value_metric ~ phenotype, scales = "free", space = "free_x") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#0072B2"),
                    name = "BH-adjusted p < 0.05") +
  labs(x = "Family effect size (eta-squared)", y = "Curve feature") +
  theme_classic(base_size = 10) +
  theme(strip.text.x = element_text(angle = 35, hjust = 0))

p_feature_rank
ggsave(file.path(fig_dir, "curve_feature_family_effect_ranking.png"),
       p_feature_rank, width = 12, height = 10)
```

```{r best-feature-boxplots, fig.width=12, fig.height=10}
best_plot_df <- curve_feature_long %>%
  semi_join(best_curve_features,
            by = c("phenotype", "value_metric", "feature")) %>%
  mutate(
    family_id_group = factor(
      family_id_group,
      levels = str_sort(unique(family_id_group), numeric = TRUE)
    ),
    panel_label = paste(phenotype, value_metric, feature, sep = "\n")
  )

family_levels <- levels(best_plot_df$family_id_group)
family_colours <- setNames(make_palette(length(family_levels)), family_levels)

p_best_features <- ggplot(best_plot_df,
    aes(x = family_id_group, y = feature_value, fill = family_id_group)) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.1) +
  facet_wrap(~ panel_label, scales = "free", ncol = 2) +
  scale_fill_manual(values = family_colours, guide = "none") +
  labs(x = "Family", y = "Feature value") +
  theme_classic(base_size = 10) +
  theme(strip.text = element_text(size = 8))

p_best_features
ggsave(file.path(fig_dir, "curve_feature_best_family_boxplots.png"),
       p_best_features, width = 12, height = 10)
```

```{r feature-heatmap, fig.width=12, fig.height=10}
heatmap_features <- curve_feature_family_stats %>%
  group_by(phenotype, value_metric) %>%
  slice_max(order_by = eta_sq, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  select(phenotype, value_metric, feature)

curve_feature_heatmap_df <- curve_feature_family_summary %>%
  semi_join(heatmap_features, by = c("phenotype", "value_metric", "feature")) %>%
  group_by(phenotype, value_metric, feature) %>%
  mutate(mean_z = as.numeric(scale(mean))) %>%
  ungroup() %>%
  mutate(
    feature_label = paste(value_metric, feature, sep = " | "),
    family_id_group = factor(
      family_id_group,
      levels = str_sort(unique(family_id_group), numeric = TRUE)
    )
  )

p_feature_heatmap <- ggplot(curve_feature_heatmap_df,
    aes(x = family_id_group, y = feature_label, fill = mean_z)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  facet_wrap(~ phenotype, scales = "free", ncol = 2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, name = "Family mean z") +
  labs(x = "Family", y = "Curve feature") +
  theme_classic(base_size = 9) +
  theme(axis.text.y = element_text(size = 6))

p_feature_heatmap
ggsave(file.path(fig_dir, "curve_feature_family_mean_heatmap.png"),
       p_feature_heatmap, width = 12, height = 10)
```

```{r trajectory-class-plot, fig.width=12, fig.height=9}
trajectory_plot_df <- trajectory_features %>%
  mutate(
    family_id_group = factor(
      family_id_group,
      levels = str_sort(unique(family_id_group), numeric = TRUE)
    )
  )

p_trajectory <- ggplot(trajectory_plot_df,
    aes(x = family_id_group, y = prop, fill = trajectory_class)) +
  geom_col(position = "stack", colour = "white", linewidth = 0.2) +
  facet_grid(value_metric ~ phenotype, scales = "free_x", space = "free_x") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Family", y = "Individuals", fill = "Trajectory class") +
  theme_classic(base_size = 9) +
  theme(strip.text.x = element_text(angle = 35, hjust = 0))

p_trajectory
ggsave(file.path(fig_dir, "curve_feature_trajectory_classes.png"),
       p_trajectory, width = 12, height = 9)
```

# Save Outputs

```{r save-outputs}
write_csv(included_experiments,
          file.path(out_dir, "included_experiments.csv"))
write_csv(experiment_overview,
          file.path(out_dir, "experiment_overview.csv"))
write_csv(curve_features,
          file.path(out_dir, "curve_features.csv"))
write_csv(curve_feature_family_summary,
          file.path(out_dir, "curve_feature_family_summary.csv"))
write_csv(curve_feature_family_stats,
          file.path(out_dir, "curve_feature_family_stats.csv"))
write_csv(curve_feature_pairwise_stats,
          file.path(out_dir, "curve_feature_pairwise_stats.csv"))
write_csv(trajectory_features,
          file.path(out_dir, "curve_feature_trajectory_classes.csv"))

message("Output files written to: ", out_dir)
```
