---
title: VIMS resazurin trials  
author: "AS Huffmyer"
date: '2025'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays conducted at VIMS.  

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
library(pracma)
library(rptR)
library(fda)
library(nls2)
library(purrr)
library(broom)
library(pheatmap)
library(vegan)
library(ggridges)
library(ggpubr)
```

# Resazurin Trials   
## Load data 

```{r}
# Set the folder path
folder_path <- "data/plate-files/"  

# List all txt files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.xlsx$", full.names = TRUE, recursive=TRUE)

file_list <- file_list[!grepl("20250630", file_list)]

# Check if any files are found
if (length(file_list) == 0) {
  stop("No xlsx files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- file_path_sans_ext(basename(file))
  
  # Read Excel file, skipping the first 9 rows
  data <- read_excel(file, skip = 10, col_names = FALSE)

  # Keep only the first 16 rows (row 1–16 after - skip)
  data <- data[1:4, ]
  
  # Assign column names: first column is row labels (A–H), others are 1–12
  colnames(data) <- c("Row", as.character(1:13))
  
  #remove first column
  #data<-data%>%select(!"Row")
  
  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Column = sprintf("%02d", as.integer(Column)),  # Ensure two-digit column numbers
      Well_ID = paste0(Row, Column),                # Format well ID as "A01", "B02", etc.
      FileName = file_name,                          # Add file name
      plate = str_extract(file_name, "plate\\d+"),  # Extract "plateX"
      date = str_extract(file_name, "^\\d{8}"),     # Extract 8-digit date
      timepoint = str_extract(file_name, "T\\d+") %>% 
        str_remove("T") %>% 
        as.numeric()                                # Convert timepoint to numeric
    ) %>%
    select(FileName, Well_ID, Value, date, plate, timepoint)  # Select relevant columns
  
  # Store the processed data in the list
  data_list[[file_name]] <- data_long
}

# Print an example of processed data
head(data_list[[file_name]])


# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  rename("well"=Well_ID, resazurin_counts=`Value`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/trial_metadata.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)

full_data<-left_join(combined_data, metadata, by=c("date", "well", "plate"))%>%
  filter(!is.na(type))

head(full_data)
```

Load in size data. 

```{r}
size<-read_xlsx("data/size/trial_size.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data. 

```{r}
str(full_data)
str(size)

full_data<-left_join(full_data, size, by=c("date", "plate", "well"))
head(full_data)

full_data<-full_data%>%
  select(!Source)%>%
  select(!FileName)%>%
  select(!notes)

head(full_data)
```

## Prep the data 

Plot the raw data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=family, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate size normalized fluorescence at each time point normalized to the starting value at time 0. 
```{r}
full_data<-full_data%>%
  group_by(date, plate, well, family)%>%
  arrange(date, plate, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=family, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour = temperature, group=interaction(date, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Calculate mean change in blank at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, plate, timepoint)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank))+
  facet_wrap(~date*plate)+
  geom_point()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(fluorescence.corr=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=family, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr, colour=length.mm, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  scale_colour_gradient(low="darkblue", high="red")+
  geom_point()+
  geom_line()+
  theme_classic()
```

Size normalize data.
```{r}
#normalize by length
full_data<-full_data%>%
  mutate(fluorescence.corr.mm=fluorescence.corr/length.mm)
```

Plot again. 

Plot by area normalized. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm, colour=family, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  theme_classic()
```

Plot by area normalized colored by size. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.corr.mm, colour=length.mm, group=interaction(date, well, plate)))+
  facet_wrap(~date)+
  scale_colour_gradient(low="darkblue", high="red")+
  geom_point()+
  geom_line()+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
str(full_data)

full_data<-full_data%>%
  #select(!resazurin_counts)%>%
  select(!type)%>%
  select(!mean_blank)
  #select(!length.mm)%>%
  #select(!fluorescence.corr)%>%
  #select(!fluorescence.norm)

str(full_data)

full_data<-full_data%>%
  rename(value=fluorescence.corr.mm)
```

### Remove oversaturated samples

Remove any sample that reached > 3000 raw value at any point during the trials, these were over saturated and data is unreliable. 

```{r}
full_data%>%
  filter(resazurin_counts>3000)%>%
  group_by(date, plate, well)%>%
  mutate(unique=paste(date, plate, well))

#there are 150 samples that went above the threshold 

list<-full_data%>%
  filter(resazurin_counts>3000)%>%
  group_by(date, plate, well)%>%
  mutate(unique=paste(date, plate, well))%>%
  pull(unique)%>%
  unique()
```

Remove the samples that exceeded the threshold of 2800 sometime during the trial. 

```{r}
full_data<-full_data%>%
  mutate(unique=paste(date, plate, well))%>%
  filter(!unique %in% list)
```

Ideal size range is in the middle range of what we used for these trials (~ >10mm and less that 18mm) 
```{r}
hist(full_data$length.mm)
```

## Separate testing and experimental plates 

Remove plates A and B from the data file and into their own data set to examine separately. 

```{r}
test_data<-full_data%>%
  filter(family %in% c("90"))

full_data<-full_data%>%
  filter(!plate %in% c("plate71", "plate72"))
```

View the test data plates. Plate A/71 was run at 4°C and plate B/72 was run at 21°C. 
```{r}
plot0<-test_data%>%
  ggplot(aes(x=timepoint, y=value, colour=temperature, group=interaction(date, family, plate, well)))+
  geom_point()+
  scale_colour_manual(labels=c("21-22°C", "35°C", "12-17°C"), values=c("orange", "darkred", "blue3"))+
  geom_line()+
  theme_classic();plot0
```

## Models and Plots - by family 

Set order of families for analysis. 

```{r}
levels_list<-c(2, 5, 6, 7, 12, 14, 15, 16, 19, 20, 21, 24, 25, 26, 31, 32, 33, 34, 36,
              37, 38, 39, 41, 42, 43, 47, 48, 49, 50, 51, 54, 55, 57, 60, 61, 70, 71,
              72, 74, 76, 77, 78, 79, 80, 82, 83, 84, 85, 88, 90)

full_data<-full_data%>%
  mutate(family=factor(family, levels=levels_list))
```

Plot individual data. 
```{r}
plot1<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(date, family, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot1

ggsave(plot1, filename="figures/trials/individual_trajectories.png", width=10, height=8, units="in")
```

Run a model on metabolic rates. 

```{r}
model<-full_data%>%
  lmer((value)^(1/3) ~ family * timepoint + (1|well:plate:date) + (1|plate) + (1|date), data=.)

anova(model)
summary(model)
qqPlot(residuals(model))
```

Plot as a model for each family at each temperature. 

```{r}
plot2<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(date, family, temperature)))+
  facet_wrap(~family)+
  geom_smooth(stat="smooth")+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2

ggsave(plot2, filename="figures/trials/family_trajectories_date.png", width=12, height=12, units="in")

plot2b<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=phenotype, group=interaction(date, family, temperature)))+
  stat_smooth(geom="smooth", method="loess", se=FALSE)+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2b

plot2c<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(family, temperature)))+
  facet_wrap(~family)+
  geom_smooth(stat="smooth")+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2c

ggsave(plot2c, filename="figures/trials/family_trajectories.png", width=13, height=12, units="in")

plot2d<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=family, group=interaction(family, temperature)))+
  geom_smooth(stat="smooth", method="loess", se=FALSE)+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2d

ggsave(plot2d, filename="figures/trials/family_trajectories2.png", width=8, height=8, units="in")
```

## Models and Plots - by phenotype 

Plot raw data. 
```{r}
plot1a<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=phenotype, group=interaction(date, phenotype, plate, well)))+
  facet_wrap(~date)+
  geom_point()+
  geom_line()+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot1a
```

Run a model on metabolic rates. 

```{r}
model<-full_data%>%
  lmer((value)^(1/3) ~ timepoint * phenotype + (1|well:plate:date) + (1|plate) + (1|date) + (1|family), data=.)

anova(model)
summary(model)
qqPlot(residuals(model))
```

Plot as a model for each phenotype. 

```{r}
plot2a<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=phenotype, group=interaction(date, phenotype)))+
  stat_smooth(geom="smooth", method="loess")+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2a

plot2b<-full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=phenotype, group=interaction(phenotype)))+
  stat_smooth(geom="smooth", method="loess")+
  geom_vline(xintercept = 3.5, linetype="dashed")+
  theme_classic();plot2b

ggsave(plot2b, filename="figures/trials/phenotype_trajectories.png", width=8, height=8, units="in")
```

## Calculate change in fluorescence between each timepoint 

Convert to wide format and calculate percent change in fluorescence.    

```{r}
data_wide <- full_data %>%
  select(date, plate, well, family, timepoint, phenotype, value)%>%
  pivot_wider(names_from = timepoint, values_from = value)

deltas <- data_wide %>%
  mutate(
    delta_T0_T1 = (`1` - `0`),
    delta_T1_T2 = (`2` - `1`)/`1`,
    delta_T2_T3 = (`3` - `2`)/`2`,
    delta_T3_T4 = (`4` - `3`)/`3`
  )
```

Convert back to long format. 

```{r}
deltas_long<-deltas%>%
  pivot_longer(names_to = "delta", values_to = "change", cols=c(delta_T0_T1:delta_T3_T4))
```

```{r}
hist(deltas_long$change)
```

Remove high and low value outliers
```{r}
deltas_long<-deltas_long%>%
  filter(change>-20 & change <40)

hist(deltas_long$change)
```

Plot deltas for each family. 

Plot all together 
```{r}
plot4<-deltas_long%>%
  ggplot(aes(x=delta, y=change, colour=family))+
  facet_wrap(~date)+
  geom_violin()+
  theme_classic();plot4

plot4a<-deltas_long%>%
  ggplot(aes(x=delta, y=change, colour=phenotype))+
  facet_wrap(~date)+
  geom_violin()+
  theme_classic();plot4a
```

T0-T1 

```{r}
plot5<-deltas%>%
  ggplot(aes(x=family, y=delta_T0_T1, colour=family, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot5

plot5a<-deltas%>%
  ggplot(aes(x=phenotype, y=delta_T0_T1, colour=phenotype, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot5a
```

T1-T2 

```{r}
plot6<-deltas%>%
  filter(delta_T1_T2>-10 & delta_T1_T2 <40)%>%
  ggplot(aes(x=family, y=delta_T1_T2, colour=family, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot6

plot6a<-deltas%>%
  filter(delta_T1_T2>-10 & delta_T1_T2 <40)%>%
  ggplot(aes(x=phenotype, y=delta_T1_T2, colour=phenotype, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot6a
```

T2-T3 

```{r}
plot6<-deltas%>%
  filter(delta_T2_T3>-10 & delta_T2_T3 <40)%>%
  ggplot(aes(x=family, y=delta_T2_T3, colour=family, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot6

plot6a<-deltas%>%
  filter(delta_T2_T3>-10 & delta_T2_T3 <40)%>%
  ggplot(aes(x=phenotype, y=delta_T2_T3, colour=phenotype, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot6a
```

T3-T4 

```{r}
plot7<-deltas%>%
  filter(delta_T3_T4>-10 & delta_T3_T4 <40)%>%
  ggplot(aes(x=family, y=delta_T3_T4, colour=family, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot7

plot7a<-deltas%>%
  filter(delta_T3_T4>-10 & delta_T3_T4 <40)%>%
  ggplot(aes(x=phenotype, y=delta_T3_T4, colour=phenotype, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot7a

plot7b<-deltas%>%
  filter(delta_T3_T4>-10 & delta_T3_T4 <40)%>%
  ggplot(aes(x=phenotype, y=delta_T3_T4, colour=phenotype, group=interaction(phenotype)))+
  geom_violin()+
  theme_classic();plot7b
```

Calculate the difference between T0-1 and T3-4 values 

```{r}
difference<-deltas%>%
  mutate(baseline=delta_T3_T4, acute=delta_T0_T1)%>%
  mutate(difference=acute-baseline)

hist(difference$difference)

plot7d<-difference%>%
  ggplot(aes(x=phenotype, y=difference, colour=phenotype, group=interaction(family)))+
  geom_violin()+
  theme_classic();plot7d

plot7e<-difference%>%
  ggplot(aes(x=phenotype, y=difference, colour=phenotype, group=interaction(phenotype)))+
  geom_violin()+
  geom_point(position=position_jitterdodge())+
  theme_classic();plot7e
```

## Calculate total change in fluorescence over the entire trial

Total fluorescence over the course of the trial is equivalent to the value at TP4 (all values start at 0).  
```{r}
end<-full_data%>%
  filter(timepoint=="4")
```

Plot total by family and phenotype. 

```{r}
plot9<-end%>%
  ggplot(aes(x=phenotype, y=value, colour=phenotype, group=interaction(date, family, temperature)))+
  facet_wrap(~date)+
  geom_violin()+
  theme_classic();plot9

ggsave(plot9, filename="figures/trials/total_values_family_date.png", width=10, height=8, units="in")

plot9b<-end%>%
  ggplot(aes(x=phenotype, y=value, colour=phenotype, group=interaction(family, temperature)))+
  geom_violin()+
  theme_classic();plot9b

ggsave(plot9b, filename="figures/trials/total_values_family.png", width=10, height=8, units="in")

plot9a<-end%>%
  ggplot(aes(x=phenotype, y=value, colour=phenotype))+
  geom_violin(aes(fill=NULL))+
  geom_point(position=position_jitterdodge())+
  theme_classic();plot9a

ggsave(plot9a, filename="figures/trials/total_values_phenotype.png", width=10, height=8, units="in")
```

```{r}
model<-end%>%
  lmer((value)^(1/3) ~ family + (1|plate:date) + (1|date), data=.)

anova(model)
summary(model)
qqPlot(residuals(model))

model<-end%>%
  lmer((value)^(1/3) ~ phenotype + (1|plate:date) + (1|date), data=.)

anova(model)
summary(model)
qqPlot(residuals(model))
```

## Area under the curve 

Area under the curve (AUC): Summarizes total metabolic activity per oyster.

```{r}
aucs <- full_data %>%
  mutate(time=as.numeric(timepoint))%>%
  group_by(date, plate, well, family, phenotype) %>%
  summarise(AUC = trapz(time, value))
```

```{r}
plot10<-ggplot(aucs, aes(x = family, y = AUC, colour = phenotype)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(title = "AUC by Family", x = "Family", y = "AUC") +
  theme_classic();plot10

ggsave(plot10, filename="figures/trials/auc_family.png", width=10, height=8, units="in")

plot10a<-ggplot(aucs, aes(x = phenotype, y = AUC, colour = phenotype)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  labs(title = "AUC by Phenotype", x = "Family", y = "AUC") +
  theme_classic(); plot10a

ggsave(plot10a, filename="figures/trials/auc_phenotype.png", width=8, height=8, units="in")
```

## Repeatability (Intra-class Correlation)

How similar are metabolic rates among oysters within the same family, compared to between families?

Close to 0 = Most variation is within families; little family structure. Metabolic rate is more influenced by individual-level noise, environment, or measurement error.

Moderate 0.3-0.5 = There's moderate consistency among family members. Families differ somewhat in metabolic rates.

High 0.6-1.0 = 	Most variation is between families; oysters from the same family have very similar metabolic rates. This suggests strong genetic or shared environmental control.
```{r}
repeat_model <- rpt(value ~ timepoint + (1|family), grname = "family", data = full_data, datatype = "Gaussian")

summary(repeat_model)
```

Repeatability (R) value is 0.135 here - most variation is within families with little family structure 

## Heritability estimates  

Heritability estimate to compute the proportion of phenotypic variance that is genetic. 
The H2 value close to 0 indicates variation is within family largely. Values between 0.2 - 0.4 indicates moderate heritability. Values >0.5 indicate stronger heritability. 

```{r}
# Convert to matrix: rows = oysters, columns = timepoints
wide_data <- full_data %>% select(!c(length.mm, fluorescence.norm, fluorescence.corr, resazurin_counts)) %>% pivot_wider(names_from = timepoint, values_from = value)

fluor_matrix <- as.matrix(wide_data[,-c(1:7)])  # remove ID columns
fluor_fd <- Data2fd(argvals = 0:4, y = t(fluor_matrix))
plot(fluor_fd)

## Estimate heritability 
anova_model <- aov(value ~ family, data = full_data)
summary(anova_model)

# Rough heritability estimate if design is appropriate:
MS_family <- summary(anova_model)[[1]]["family", "Mean Sq"]
MS_resid <- summary(anova_model)[[1]]["Residuals", "Mean Sq"]
k<-length(unique(full_data$unique))

heritability <- (MS_family - MS_resid) / (MS_family + (k - 1) * MS_resid)  # k = number of replicates
heritability
```

Heritability = 0.0074 (low heritability, matching family variation above).  

## Fit nonlinear curves 

```{r}
# Example for one oyster
oyster_data <- full_data %>% filter(unique == "20250701 plate1 A02")%>%
  mutate(time=as.numeric(timepoint))

fit <- nls(value ~ a * exp(b * time), data = oyster_data,
           start = list(a = 100, b = -0.1))

summary(fit)

oyster_data <- oyster_data %>%
  mutate(predicted = predict(fit, newdata = oyster_data))

ggplot(oyster_data, aes(x = time, y = value)) +
  geom_point(color = "blue", size = 3) +  # Raw data points
  geom_line(aes(y = predicted), color = "red", linewidth = 1.2) +  # Fitted curve
  labs(title = "Fitted Exponential Curve vs. Raw Data",
       x = "Time (hours)",
       y = "Fluorescence (Metabolic Rate)") +
  theme_minimal()
```

Run for all oysters 

```{r}
model_results <- full_data %>%
  mutate(time = as.numeric(timepoint)) %>%
  group_by(unique, family, phenotype) %>%
  nest() %>%
  mutate(
    fit = map(data, ~ tryCatch(
      nls(value ~ a * exp(b * time), data = .x,
          start = list(a = 100, b = -0.1),
          control = nls.control(maxiter = 100)),
      error = function(e) NULL
    )),
    params = map(fit, ~ if (!is.null(.x)) tidy(.x) else NULL)
  ) %>%
  unnest(params) %>%
  select(unique, family, phenotype, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate)

family_summary <- model_results %>%
  group_by(family, phenotype) %>%
  summarise(
    mean_a = mean(a, na.rm = TRUE),
    sd_a = sd(a, na.rm = TRUE),
    mean_b = mean(b, na.rm = TRUE),
    sd_b = sd(b, na.rm = TRUE),
    n = n()
  )

plot11<-ggplot(family_summary, aes(x = reorder(family, mean_b), y = mean_b)) +
  facet_wrap(~phenotype)+
  geom_point(color = "darkred", size = 3) +
  geom_errorbar(aes(ymin = mean_b - sd_b, ymax = mean_b + sd_b), width = 0.2) +
  labs(title = "Average Exponential Decay Rate (b) by Family",
       x = "Family", y = "Mean b (Rate Constant)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1));plot11

ggsave(plot11, filename="figures/trials/exponential_growth_family.png", width=12, height=8, units="in")
```

Rate of metabolic growth/metabolic change over time is positive for all and is higher for some families.  

## Clustering metabolic profiles 

Cluster to find groups of oysters that behave similarly.  

```{r}
df_matrix <- full_data %>%
  select(!c(resazurin_counts, length.mm, fluorescence.norm, fluorescence.corr))%>%
  pivot_wider(names_from = timepoint, values_from = value) %>%
  ungroup()%>%
  select(!c(well, date, plate, family, temperature, phenotype))%>%
  column_to_rownames("unique") %>%
  as.matrix()

clusters <- kmeans(df_matrix, centers = 3)

df_clustered <- data.frame(oyster_id = rownames(df_matrix), cluster = clusters$cluster)
```

Plot a heatmap of cluster. 
```{r}
# Reorder matrix by cluster assignment
ordered_matrix <- df_matrix[order(clusters$cluster), ]

# Annotate rows with cluster info
annotation <- data.frame(cluster = as.factor(clusters$cluster))
rownames(annotation) <- rownames(df_matrix)

# Plot heatmap
pheatmap(ordered_matrix, cluster_rows = FALSE, cluster_cols = FALSE,
         annotation_row = annotation,
         main = "Heatmap of Fluorescence Profiles by Cluster")
```

Plot curves by cluster 
```{r}
# Reshape df_matrix back into long format for plotting
df_long <- df_matrix %>%
  as.data.frame() %>%
  rownames_to_column("oyster_id") %>%
  pivot_longer(-oyster_id, names_to = "timepoint", values_to = "value") %>%
  left_join(df_clustered, by = c("oyster_id")) %>%
  mutate(time = as.numeric(gsub("T", "", timepoint)))  # convert T0, T1, etc. to numeric

# Plot average fluorescence over time by cluster
plot12<-ggplot(df_long, aes(x = time, y = value, group = oyster_id)) +
  stat_summary(aes(color = as.factor(cluster)), fun = mean, geom = "line", size = 1.2) +
  labs(title = "Average Fluorescence Curve by Cluster",
       x = "Time (hours)", y = "Fluorescence", color = "Cluster") +
  theme_classic();plot12

ggsave(plot12, filename="figures/trials/cluster_fluorescence_patterns.png", width=8, height=8, units="in")
```

Show families. 

```{r}
df_clustered_full <- full_data %>%
  select(unique, family) %>%
  distinct() %>%
  left_join(df_clustered, by = c("unique" = "oyster_id"))

plot13<-ggplot(df_clustered_full, aes(x = family, fill = factor(cluster))) +
  geom_bar(position = "fill") +  # stacked proportional bar
  labs(title = "Proportion of Clusters Within Each Family",
       x = "Family", y = "Proportion of Oysters", fill = "Cluster") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  theme_classic();plot13

ggsave(plot13, filename="figures/trials/cluster_assignments_family.png", width=10, height=8, units="in")
```

## Ridge plots 

View distrubution of fluorescence values over time. 

```{r}
plot14<-ggplot(full_data, aes(x = value, y = as.factor(timepoint), fill = family)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Distribution of Fluorescence by Family")+
  theme_classic();plot14

ggsave(plot14, filename="figures/trials/ridge_family.png", width=10, height=8, units="in")

plot15<-ggplot(full_data, aes(x = value, y = as.factor(timepoint), fill = phenotype)) +
  geom_density_ridges(alpha = 0.7) +
  labs(title = "Distribution of Fluorescence by Phenotype")+
  theme_classic();plot15

ggsave(plot15, filename="figures/trials/ridge_phenotype.png", width=10, height=8, units="in")
```

## Correlation with traits 

Load in predicted performance data provided by the breeding program. Data are recorded as percent (proportion) difference in a trait from the general population mean. For example, 0.50 indicates a 50% higher value than general population.  

```{r}
performance<-read_excel(path="data/predicted_performance.xlsx")%>%mutate(family=as.factor(family))

str(performance)
```

### Correlate with total metabolic activity (AUC)

Merge into the AUC data set to correlate with total metabolic rates.

```{r}
merged<-left_join(aucs, performance)
```

Correlation of survival in high salinity and total metabolic activity. 

```{r}
plot15<-merged%>%
  group_by(family, phenotype, `survival-high-salinity`)%>%
  summarise(AUC=mean(AUC))%>%
  
  ggplot(., aes(x = AUC, y = `survival-high-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged$AUC), label.y = max(merged$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Total Metabolic Activity (AUC)",
    y = "Survival in High Salinity"
  )+
  theme_classic();plot15

ggsave(plot15, filename="figures/correlations/auc_survival-high-salinity.png", width=8, height=6, units="in")
```

Correlation of survival in low salinity and total metabolic activity. 

```{r}
plot16<-merged%>%
  group_by(family, phenotype, `survival-low-salinity`)%>%
  summarise(AUC=mean(AUC))%>%
  
  ggplot(., aes(x = AUC, y = `survival-low-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged$AUC), label.y = max(merged$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Total Metabolic Activity (AUC)",
    y = "Survival in Low Salinity"
  )+
  theme_classic();plot16

ggsave(plot16, filename="figures/correlations/auc_survival-low-salinity.png", width=8, height=6, units="in")
```

Correlation of weight in high salinity and total metabolic activity. 

```{r}
plot17<-merged%>%
  group_by(family, phenotype, `weight-high-salinity`)%>%
  summarise(AUC=mean(AUC))%>%
  
  ggplot(., aes(x = AUC, y = `weight-high-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged$AUC), label.y = max(merged$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Total Metabolic Activity (AUC)",
    y = "Weight in High Salinity"
  )+
  theme_classic();plot17

ggsave(plot17, filename="figures/correlations/auc_weight-high-salinity.png", width=8, height=6, units="in")
```

Correlation of weight in low salinity and total metabolic activity. 
```{r}
plot18<-merged%>%
  group_by(family, phenotype, `weight-low-salinity`)%>%
  summarise(AUC=mean(AUC))%>%
  
  ggplot(., aes(x = AUC, y = `weight-low-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged$AUC), label.y = max(merged$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Total Metabolic Activity (AUC)",
    y = "Weight in Low Salinity"
  )+
  theme_classic();plot18

ggsave(plot18, filename="figures/correlations/auc_weight-low-salinity.png", width=8, height=6, units="in")
```

### Correlate with metabolic slope in response to acute stress (Delta T0-T1)

```{r}
str(deltas)

merged2<-left_join(deltas, performance)
```

Correlation of survival in high salinity and total metabolic activity. 

```{r}
plot19<-merged2%>%
  group_by(family, phenotype, `survival-high-salinity`)%>%
  summarise(delta=mean(delta_T0_T1))%>%
  
  ggplot(., aes(x = delta, y = `survival-high-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged2$delta), label.y = max(merged2$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Metabolic Rate - Acute Stress",
    y = "Survival in High Salinity"
  )+
  theme_classic();plot19

ggsave(plot19, filename="figures/correlations/stress-rate_survival-high-salinity.png", width=8, height=6, units="in")
```

Correlation of survival in low salinity and total metabolic activity. 

```{r}
plot20<-merged2%>%
  group_by(family, phenotype, `survival-low-salinity`)%>%
  summarise(delta=mean(delta_T0_T1))%>%
  
  ggplot(., aes(x = delta, y = `survival-low-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged2$delta), label.y = max(merged2$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Metabolic Rate - Acute Stress",
    y = "Survival in Low Salinity"
  )+
  theme_classic();plot20

ggsave(plot20, filename="figures/correlations/stress-rate_survival-low-salinity.png", width=8, height=6, units="in")
```

Correlation of weight in high salinity and total metabolic activity. 

```{r}
plot21<-merged2%>%
  group_by(family, phenotype, `weight-high-salinity`)%>%
  summarise(delta=mean(delta_T0_T1))%>%
  
  ggplot(., aes(x = delta, y = `weight-high-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged2$delta), label.y = max(merged2$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Metabolic Rate - Acute Stress",
    y = "Weight in High Salinity"
  )+
  theme_classic();plot21

ggsave(plot21, filename="figures/correlations/stress-rate_weight-high-salinity.png", width=8, height=6, units="in")
```

Correlation of weight in low salinity and total metabolic activity. 
```{r}
plot22<-merged2%>%
  group_by(family, phenotype, `weight-low-salinity`)%>%
  summarise(delta=mean(delta_T0_T1))%>%
  
  ggplot(., aes(x = delta, y = `weight-low-salinity`)) +
  facet_wrap(~phenotype)+
  geom_point(color = "steelblue", size = 2) +  # scatter points
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linetype = "dashed") +  # regression line
  stat_cor(method = "pearson", label.x = min(merged2$delta), label.y = max(merged2$`survival-high-salinity`), 
           aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), 
           size = 5) +  # correlation stat
  theme_minimal() +
  labs(
    title = "Scatter Plot with Pearson Correlation",
    x = "Total Metabolic Activity (AUC)",
    y = "Weight in Low Salinity"
  )+
  theme_classic();plot22

ggsave(plot22, filename="figures/correlations/stress-rate_weight-low-salinity.png", width=8, height=6, units="in")
```

