---
title: "resazurin-family-phenotype-prediction"
author: "Ariana Huffmyer"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options:
  chunk_output_type: console
---

# Background

**Question:** *How well can a resazurin metabolic assay predict the survival
phenotype of an oyster family, and do the same metabolic features predict
performance under high-salinity vs low-salinity conditions?*

Two independent datasets describe the same 48 families:

1. **Survival phenotypes** — `data/predicted_performance.xlsx` records each
   family's observed survival proportion under both high-salinity
   (`survival-high-salinity`) and low-salinity (`survival-low-salinity`)
   conditions, along with family rankings within each condition
   (`rank-survival-high-salinity`, `rank-survival-low-salinity`; 1 = best).
   Families span two parentage groups: high-salinity and low-salinity origin.

2. **Metabolic curve features** — `output/resazurin-curves/curve_features.csv`
   contains ~25 curve shape/capacity/rate traits per individual extracted from
   resazurin assay timecourses (AUC, Vmax, metabolic scope, depression index,
   resilience ratio, trajectory class, etc.).

We ask (a) which resazurin curve features predict survival under each salinity
condition, and (b) whether different features predict performance under the two
conditions, which would reveal condition-specific metabolic signals.

## Expected inputs

| Path | Description |
|------|-------------|
| `output/resazurin-curves/curve_features.csv` | Per-individual curve traits |
| `data/predicted_performance.xlsx` | Family survival proportions and rankings |

## Expected outputs

All written to `output/resazurin-prediction/`.

| File | Description |
|------|-------------|
| `family_feature_matrix.csv` | Family-mean resazurin features by parentage group |
| `feature_phenotype_correlations.csv` | Spearman/Pearson correlations for all features × phenotypes |
| `top_predictors.csv` | Best-correlated features for each survival condition |
| `lofo_cv_performance.csv` | LOFO-CV performance for candidate predictors |
| `resazurin_index_predictions.csv` | Per-family observed vs predicted for composite index |
| `figures/` | Correlation ranking, comparison scatter, best-predictor, index, heatmap |

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  results = "hold"
)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(here)
library(ggrepel)
library(cowplot)
```

```{r paths}
resazurin_features_path <- here("output", "resazurin-curves", "curve_features.csv")
phenotype_path          <- here("data", "predicted_performance.xlsx")

out_dir <- here("output", "resazurin-prediction")
fig_dir <- file.path(out_dir, "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(resazurin_features_path), file.exists(phenotype_path))

# The two primary survival conditions analysed in parallel
PRIMARY_PHENOTYPES <- c("survival_high_salinity", "survival_low_salinity")
PHENOTYPE_LABELS   <- c(
  survival_high_salinity = "High-salinity survival",
  survival_low_salinity  = "Low-salinity survival"
)
```

# Load the survival phenotype

Both survival proportions (0–1, higher = better) are the primary phenotypes.
Family ranks (1 = best within condition) are retained as secondary variables.

```{r load-phenotype}
phenotype <- read_excel(phenotype_path) %>%
  transmute(
    family                       = as.character(family),
    survival_high_salinity       = `survival-high-salinity`,
    survival_low_salinity        = `survival-low-salinity`,
    rank_survival_high_salinity  = `rank-survival-high-salinity`,
    rank_survival_low_salinity   = `rank-survival-low-salinity`
  )

cat("Families with phenotype data:", nrow(phenotype), "\n")
cat("High-salinity survival range:",
    round(range(phenotype$survival_high_salinity), 3), "\n")
cat("Low-salinity survival range:",
    round(range(phenotype$survival_low_salinity), 3), "\n")
phenotype
```

# Load resazurin curve features

```{r load-resazurin}
resazurin_raw <- read_csv(resazurin_features_path, show_col_types = FALSE) %>%
  rename(parentage_group = phenotype) %>%
  mutate(family = str_to_lower(trimws(as.character(family_id_group))))

resazurin <- resazurin_raw %>%
  filter(family %in% phenotype$family)

cat("Families in resazurin data:    ", n_distinct(resazurin_raw$family), "\n")
cat("Families matched to phenotype: ", n_distinct(resazurin$family), "\n")
cat("Parentage groups:              ",
    paste(sort(unique(resazurin$parentage_group)), collapse = ", "), "\n")
cat("Value metrics:                 ",
    paste(sort(unique(resazurin$value_metric)), collapse = ", "), "\n")
```

# Build the family-level feature matrix

```{r feature-matrix}
feature_cols <- resazurin %>%
  select(where(is.numeric)) %>%
  select(-any_of(c("n_timepoints", "family_id_group"))) %>%
  names()
feature_cols <- setdiff(feature_cols, "baseline_value")

family_feature_long <- resazurin %>%
  select(parentage_group, value_metric, family, all_of(feature_cols)) %>%
  pivot_longer(all_of(feature_cols),
               names_to = "feature", values_to = "value") %>%
  filter(is.finite(value)) %>%
  bind_rows({.} %>% mutate(parentage_group = "all")) %>%
  group_by(parentage_group, value_metric, feature, family) %>%
  summarise(
    n_indiv     = n(),
    family_mean = mean(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(parentage_group, value_metric, feature) %>%
  filter(n_distinct(family) >= 7) %>%
  ungroup()

write_csv(family_feature_long, file.path(out_dir, "family_feature_matrix.csv"))

family_feature_long %>%
  count(parentage_group, value_metric, name = "n_feature_rows") %>%
  arrange(parentage_group, value_metric)
```

# Screen: correlate each feature with survival phenotypes

Spearman `rho` (rank agreement) is primary; Pearson `r` is reported alongside.
Both survival conditions are assessed in parallel.

```{r correlations}
phenotypes_long <- phenotype %>%
  select(family, all_of(PRIMARY_PHENOTYPES)) %>%
  pivot_longer(-family, names_to = "survival_condition",
               values_to = "pheno_value")

cor_one <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 7 || sd(x[ok]) == 0 || sd(y[ok]) == 0) {
    return(tibble(n = sum(ok), spearman_rho = NA_real_, spearman_p = NA_real_,
                  pearson_r = NA_real_, pearson_p = NA_real_))
  }
  sp <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman"))
  pe <- suppressWarnings(cor.test(x[ok], y[ok], method = "pearson"))
  tibble(n = sum(ok),
         spearman_rho = unname(sp$estimate), spearman_p = sp$p.value,
         pearson_r    = unname(pe$estimate), pearson_p  = pe$p.value)
}

feature_phenotype_correlations <- family_feature_long %>%
  inner_join(phenotypes_long, by = "family",
             relationship = "many-to-many") %>%
  group_by(parentage_group, value_metric, feature, survival_condition) %>%
  group_modify(~ cor_one(.x$family_mean, .x$pheno_value)) %>%
  ungroup() %>%
  arrange(survival_condition, desc(abs(spearman_rho)))

write_csv(feature_phenotype_correlations,
          file.path(out_dir, "feature_phenotype_correlations.csv"))

# Top predictors for each survival condition
top_predictors <- feature_phenotype_correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(survival_condition) %>%
  arrange(desc(abs(spearman_rho))) %>%
  slice_head(n = 15) %>%
  ungroup()

write_csv(top_predictors, file.path(out_dir, "top_predictors.csv"))

knitr::kable(
  top_predictors %>%
    transmute(survival_condition, parentage_group, feature, n,
              spearman_rho = round(spearman_rho, 3),
              spearman_p   = signif(spearman_p, 3),
              pearson_r    = round(pearson_r, 3)),
  caption = "Top resazurin features predicting survival under each salinity condition"
)
```

```{r cor-ranking-plot, fig.width=12, fig.height=8}
plot_cor <- feature_phenotype_correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(survival_condition, parentage_group, value_metric) %>%
  slice_max(abs(spearman_rho), n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    survival_label = PHENOTYPE_LABELS[survival_condition],
    feature        = reorder(feature, abs(spearman_rho))
  )

p_cor <- ggplot(plot_cor,
    aes(x = spearman_rho, y = feature, fill = spearman_p < 0.05)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  facet_grid(survival_label ~ parentage_group, scales = "free_y") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#0072B2"),
                    name = "Spearman p < 0.05") +
  labs(x = "Spearman rho with survival proportion",
       y = "Resazurin curve feature",
       title = "Resazurin features vs survival under each salinity condition") +
  theme_classic(base_size = 10) +
  theme(strip.text = element_text(size = 9))

p_cor
ggsave(file.path(fig_dir, "feature_phenotype_correlation_ranking.png"),
       p_cor, width = 12, height = 8)
```

# Comparing predictors: high-salinity vs low-salinity survival

How does each feature's predictive power for high-salinity survival compare to
its power for low-salinity survival? Features that fall off the diagonal predict
one condition much better than the other.

```{r feature-comparison, fig.width=9, fig.height=7}
comparison_df <- feature_phenotype_correlations %>%
  filter(is.finite(spearman_rho)) %>%
  select(parentage_group, value_metric, feature, survival_condition, spearman_rho) %>%
  pivot_wider(names_from  = survival_condition,
              values_from = spearman_rho) %>%
  filter(is.finite(survival_high_salinity), is.finite(survival_low_salinity)) %>%
  mutate(rho_diff      = survival_high_salinity - survival_low_salinity,
         abs_rho_max   = pmax(abs(survival_high_salinity),
                              abs(survival_low_salinity)),
         label_feature = if_else(abs_rho_max > 0.25 | abs(rho_diff) > 0.15,
                                 feature, NA_character_))

p_compare <- ggplot(comparison_df,
    aes(x = survival_high_salinity, y = survival_low_salinity,
        colour = parentage_group, shape = parentage_group)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.25, colour = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.25, colour = "grey70") +
  geom_point(size = 2.5, alpha = 0.8) +
  ggrepel::geom_text_repel(aes(label = label_feature),
                            size = 3, max.overlaps = 15, show.legend = FALSE,
                            na.rm = TRUE) +
  scale_colour_manual(values = c("all"           = "#444444",
                                 "high-salinity" = "#D55E00",
                                 "low-salinity"  = "#0072B2")) +
  scale_shape_manual(values = c("all" = 15, "high-salinity" = 16,
                                "low-salinity" = 17)) +
  labs(
    x      = "Spearman rho — high-salinity survival",
    y      = "Spearman rho — low-salinity survival",
    colour = "Parentage group", shape = "Parentage group",
    title  = "Feature predictive power: high-salinity vs low-salinity survival",
    subtitle = "Dashed line = equal prediction; points above = better low-sal predictor"
  ) +
  theme_classic(base_size = 11)

p_compare
ggsave(file.path(fig_dir, "feature_comparison_high_vs_low_salinity.png"),
       p_compare, width = 9, height = 7)

# Table: features with the largest divergence in predictive direction
cat("\nFeatures with largest differential predictive power",
    "(|rho_high - rho_low| > 0.10):\n\n")
comparison_df %>%
  filter(abs(rho_diff) > 0.10) %>%
  arrange(desc(abs(rho_diff))) %>%
  transmute(parentage_group, feature,
            rho_high_sal = round(survival_high_salinity, 3),
            rho_low_sal  = round(survival_low_salinity,  3),
            rho_diff     = round(rho_diff, 3)) %>%
  print(n = 20)
```

# Cross-validated prediction (leave-one-family-out)

```{r lofo-fn}
lofo_cv <- function(df) {
  df <- df %>% filter(is.finite(x), is.finite(y))
  if (nrow(df) < 7 || sd(df$x) == 0) {
    return(tibble(n = nrow(df), cv_spearman = NA_real_, cv_rmse = NA_real_))
  }
  preds <- map_dbl(seq_len(nrow(df)), function(i) {
    fit <- lm(y ~ x, data = df[-i, ])
    predict(fit, newdata = df[i, ])
  })
  sp <- suppressWarnings(cor(preds, df$y, method = "spearman"))
  tibble(n = nrow(df), cv_spearman = sp,
         cv_rmse = sqrt(mean((preds - df$y)^2)))
}
```

```{r lofo-candidates}
# Top 3 candidate features per condition × parentage_group × metric
candidates <- feature_phenotype_correlations %>%
  filter(is.finite(spearman_rho), n >= 7) %>%
  group_by(survival_condition, parentage_group, value_metric) %>%
  slice_max(abs(spearman_rho), n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  select(survival_condition, parentage_group, value_metric, feature)

lofo_cv_performance <- candidates %>%
  left_join(family_feature_long,
            by = c("parentage_group", "value_metric", "feature"),
            relationship = "many-to-many") %>%
  left_join(
    phenotypes_long %>% rename(survival_condition = survival_condition),
    by = c("family", "survival_condition")
  ) %>%
  rename(x = family_mean, y = pheno_value) %>%
  group_by(survival_condition, parentage_group, value_metric, feature) %>%
  group_modify(~ lofo_cv(.x)) %>%
  ungroup() %>%
  left_join(
    feature_phenotype_correlations %>%
      select(survival_condition, parentage_group, value_metric,
             feature, spearman_rho),
    by = c("survival_condition", "parentage_group", "value_metric", "feature")
  ) %>%
  arrange(survival_condition, desc(cv_spearman))

write_csv(lofo_cv_performance,
          file.path(out_dir, "lofo_cv_performance.csv"))

knitr::kable(
  lofo_cv_performance %>%
    transmute(survival_condition, parentage_group, feature, n,
              in_sample_rho = round(spearman_rho, 3),
              cv_spearman   = round(cv_spearman,  3),
              cv_rmse       = round(cv_rmse, 4)),
  caption = "Leave-one-family-out CV: single-feature prediction of survival"
)
```

## Best single predictor per condition

```{r best-single-scatter, fig.width=14, fig.height=6}
best_singles <- lofo_cv_performance %>%
  filter(is.finite(cv_spearman)) %>%
  group_by(survival_condition) %>%
  slice_max(cv_spearman, n = 1, with_ties = FALSE) %>%
  ungroup()

plots_best <- map(PRIMARY_PHENOTYPES, function(cond) {
  bs  <- best_singles %>% filter(survival_condition == cond)
  df  <- family_feature_long %>%
    semi_join(bs, by = c("parentage_group", "value_metric", "feature")) %>%
    left_join(phenotype, by = "family") %>%
    mutate(pheno_val = .data[[cond]])

  ggplot(df, aes(x = family_mean, y = pheno_val)) +
    geom_smooth(method = "lm", se = TRUE, color = "#0072B2",
                fill = "#0072B2", alpha = 0.15) +
    geom_point(size = 2.5) +
    geom_text(aes(label = family), size = 3, vjust = -0.8) +
    labs(
      title    = PHENOTYPE_LABELS[cond],
      subtitle = sprintf("Best predictor: %s (%s) | rho = %.2f | CV rho = %.2f",
                         bs$feature, bs$parentage_group,
                         bs$spearman_rho, bs$cv_spearman),
      x = sprintf("Family-mean %s", bs$feature),
      y = PHENOTYPE_LABELS[cond]
    ) +
    theme_classic(base_size = 11)
})

p_best_both <- cowplot::plot_grid(plotlist = plots_best, nrow = 1,
                                  labels = "AUTO")
p_best_both
ggsave(file.path(fig_dir, "best_single_predictor_scatter.png"),
       p_best_both, width = 14, height = 6)
```

# Multi-feature resazurin index (per survival condition)

For each survival condition, a simple index is built from the top 4 features
in the best-performing parentage group: z-score each, sign-align so that
higher always means "more survival-like", then average.

```{r resazurin-index}
build_index <- function(cond) {
  best_grp <- lofo_cv_performance %>%
    filter(survival_condition == cond, is.finite(cv_spearman)) %>%
    slice_max(cv_spearman, n = 1, with_ties = FALSE)

  idx_feats <- feature_phenotype_correlations %>%
    filter(survival_condition == cond,
           parentage_group == best_grp$parentage_group,
           value_metric    == best_grp$value_metric,
           is.finite(spearman_rho), n >= 7) %>%
    arrange(desc(abs(spearman_rho))) %>%
    slice_head(n = 4)

  cat("\n---", PHENOTYPE_LABELS[cond], "---\n")
  cat("Best group:", best_grp$parentage_group,
      "| metric:", best_grp$value_metric, "\n")
  cat("Index features:\n")
  idx_feats %>%
    transmute(feature, spearman_rho = round(spearman_rho, 3)) %>%
    as.data.frame() %>% print(row.names = FALSE)

  wide <- family_feature_long %>%
    filter(parentage_group == best_grp$parentage_group,
           value_metric    == best_grp$value_metric) %>%
    semi_join(idx_feats, by = c("parentage_group", "value_metric", "feature")) %>%
    left_join(idx_feats %>% select(feature, spearman_rho), by = "feature") %>%
    group_by(feature) %>%
    mutate(z = as.numeric(scale(family_mean)),
           z_signed = z * sign(spearman_rho)) %>%
    ungroup() %>%
    group_by(family) %>%
    summarise(resazurin_index = mean(z_signed, na.rm = TRUE), .groups = "drop") %>%
    left_join(phenotype, by = "family") %>%
    mutate(survival_condition   = cond,
           parentage_group_used = best_grp$parentage_group,
           pheno_val            = .data[[cond]])

  idx_cv <- lofo_cv(wide %>%
                      transmute(family, x = resazurin_index, y = pheno_val))

  fit <- lm(pheno_val ~ resazurin_index, data = wide)
  wide <- wide %>%
    mutate(predicted_survival = as.numeric(predict(fit)))

  cat(sprintf("Index rho = %.2f | LOFO-CV rho = %.2f | CV RMSE = %.4f\n",
              suppressWarnings(
                cor(wide$resazurin_index, wide$pheno_val,
                    method = "spearman")),
              idx_cv$cv_spearman, idx_cv$cv_rmse))

  list(wide = wide, cv = idx_cv,
       best_grp = best_grp$parentage_group,
       best_metric = best_grp$value_metric)
}

idx_results <- map(PRIMARY_PHENOTYPES, build_index)
names(idx_results) <- PRIMARY_PHENOTYPES

index_all <- map_dfr(idx_results, ~ .x$wide)
write_csv(index_all, file.path(out_dir, "resazurin_index_predictions.csv"))
```

```{r index-scatter, fig.width=14, fig.height=6}
plots_idx <- imap(idx_results, function(res, cond) {
  ggplot(res$wide, aes(x = resazurin_index, y = pheno_val)) +
    geom_smooth(method = "lm", se = TRUE, color = "#009E73",
                fill = "#009E73", alpha = 0.15) +
    geom_point(size = 2.5) +
    geom_text(aes(label = family), size = 3, vjust = -0.8) +
    labs(
      title    = PHENOTYPE_LABELS[cond],
      subtitle = sprintf("group = %s | LOFO-CV rho = %.2f",
                         res$best_grp, res$cv$cv_spearman),
      x = "Resazurin index (higher = more survival-like)",
      y = PHENOTYPE_LABELS[cond]
    ) +
    theme_classic(base_size = 11)
})

p_idx_both <- cowplot::plot_grid(plotlist = plots_idx, nrow = 1,
                                 labels = "AUTO")
p_idx_both
ggsave(file.path(fig_dir, "resazurin_index_scatter.png"),
       p_idx_both, width = 14, height = 6)
```

# Feature heatmaps ordered by survival rank

Families are ordered from highest to lowest survival within each condition.

```{r profile-heatmap, fig.width=13, fig.height=8}
plots_heat <- imap(idx_results, function(res, cond) {
  rank_col <- if (cond == "survival_high_salinity") {
    "rank_survival_high_salinity"
  } else {
    "rank_survival_low_salinity"
  }
  rank_order <- phenotype %>%
    arrange(.data[[rank_col]]) %>%
    pull(family)

  heat_features <- feature_phenotype_correlations %>%
    filter(survival_condition == cond,
           parentage_group    == res$best_grp,
           value_metric       == res$best_metric,
           is.finite(spearman_rho)) %>%
    slice_max(abs(spearman_rho), n = 10, with_ties = FALSE) %>%
    pull(feature)

  heat_df <- family_feature_long %>%
    filter(parentage_group == res$best_grp,
           value_metric    == res$best_metric,
           feature         %in% heat_features) %>%
    group_by(feature) %>%
    mutate(z = as.numeric(scale(family_mean))) %>%
    ungroup() %>%
    mutate(family = factor(family, levels = rank_order))

  ggplot(heat_df, aes(x = family, y = feature, fill = z)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, name = "z-score") +
    labs(
      title    = PHENOTYPE_LABELS[cond],
      subtitle = sprintf("group = %s  (left = best survivor)",
                         res$best_grp),
      x = "Family (highest → lowest survival)",
      y = "Curve feature"
    ) +
    theme_classic(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
})

p_heat_both <- cowplot::plot_grid(plotlist = plots_heat, nrow = 1,
                                  labels = "AUTO")
p_heat_both
ggsave(file.path(fig_dir, "family_profile_heatmap.png"),
       p_heat_both, width = 13, height = 8)
```

# Synthesis

```{r synthesis, results='asis'}
for (cond in PRIMARY_PHENOTYPES) {
  cat("##", PHENOTYPE_LABELS[cond], "\n\n")

  top3 <- feature_phenotype_correlations %>%
    filter(survival_condition == cond, is.finite(spearman_rho)) %>%
    arrange(desc(abs(spearman_rho))) %>%
    slice_head(n = 3)

  best_cv3 <- lofo_cv_performance %>%
    filter(survival_condition == cond, is.finite(cv_spearman)) %>%
    arrange(desc(cv_spearman)) %>%
    slice_head(n = 3)

  cat("**Strongest raw associations:**\n\n")
  for (i in seq_len(nrow(top3))) {
    cat(sprintf("- **%s** (%s / %s): rho = %.2f (p = %.3g)\n",
                top3$feature[i], top3$value_metric[i],
                top3$parentage_group[i],
                top3$spearman_rho[i], top3$spearman_p[i]))
  }

  cat("\n**Best LOFO-CV single predictors:**\n\n")
  for (i in seq_len(nrow(best_cv3))) {
    cat(sprintf("- **%s** (%s / %s): CV rho = %.2f, CV RMSE = %.4f\n",
                best_cv3$feature[i], best_cv3$value_metric[i],
                best_cv3$parentage_group[i],
                best_cv3$cv_spearman[i], best_cv3$cv_rmse[i]))
  }

  res <- idx_results[[cond]]
  cat(sprintf("\nMulti-feature index (group = %s): LOFO-CV rho = %.2f\n\n",
              res$best_grp, res$cv$cv_spearman))
}

cat("## High-salinity vs low-salinity comparison\n\n")
div_feats <- comparison_df %>%
  filter(abs(rho_diff) > 0.10) %>%
  arrange(desc(abs(rho_diff))) %>%
  slice_head(n = 5)

cat("Features with most divergent predictive power across salinity conditions:\n\n")
for (i in seq_len(nrow(div_feats))) {
  direction <- if (div_feats$rho_diff[i] > 0) {
    "stronger for high-salinity survival"
  } else {
    "stronger for low-salinity survival"
  }
  cat(sprintf("- **%s** (%s): rho_high = %.2f, rho_low = %.2f — %s\n",
              div_feats$feature[i], div_feats$parentage_group[i],
              div_feats$survival_high_salinity[i],
              div_feats$survival_low_salinity[i],
              direction))
}
```

# Session info

```{r session-info}
sessionInfo()
```
