---
title: "resazurin-ploidy-field-phenotype-prediction"
author: "Ariana Huffmyer"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  github_document:
    toc: true
    number_sections: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options:
  chunk_output_type: console
---

# Background

Two independent datasets describe the same set of diploid and triploid oysters:

1. **Field phenotypes** — `data/ploidy/field/ploidy_field_performance.xlsx` records each
   ploidy's observed performance. 
   
- Definitions: 

    - SH is shell height (longest measure)
    - SL is shell length (across the ‘fan’)
    - SW is shell width (typically the shortest dimension)
    - Fan is SL/SH
    - Cup is SW/SH
    - WWW is whole wet weight
    - DSW is dry shell weight
    - DTW is dry tissue weight
    - CI is condition index and we use 100 x DTW/(WWW-DSW)
    - Live dead is survival 

2. **Metabolic curve features** — `output/ploidy/resazurin-curves/curve_features.csv`
   contains ~25 curve shape/capacity/rate traits per individual extracted from
   resazurin assay timecourses (AUC, Vmax, metabolic scope, depression index,
   resilience ratio, trajectory class, etc.).

## Expected inputs

| Path | Description |
|------|-------------|
| `output/ploidy/resazurin-curves/curve_features.csv` | Per-individual curve traits |
| `data/ploidy/field/ploidy/field_performance.xlsx` | Ploidy performance |

## Expected outputs

All written to `output/ploidy/resazurin-prediction/`.

| File | Description |
|------|-------------|
| `family_feature_matrix.csv` | Ploidy-mean resazurin features by parentage group |
| `feature_phenotype_correlations.csv` | Spearman/Pearson correlations for all features |
| `top_predictors.csv` | Best-correlated features for each survival condition |
| `lofo_cv_performance.csv` | LOFO-CV performance for candidate predictors |
| `resazurin_index_predictions.csv` | Per-family observed vs predicted for composite index |
| `figures/` | Correlation ranking, comparison scatter, best-predictor, index, heatmap |

# Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  results = "hold"
)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(here)
library(ggrepel)
library(cowplot)
```

```{r paths}
resazurin_features_path <- here("output/ploidy", "resazurin-curves", "curve_features.csv")
phenotype_path          <- here("data/ploidy/field", "ploidy_field_performance.xlsx")

out_dir <- here("output/ploidy", "resazurin-prediction")
fig_dir <- file.path(out_dir, "figures/ploidy")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

stopifnot(file.exists(resazurin_features_path), file.exists(phenotype_path))
```

# Load the survival phenotype

```{r load-phenotype}
phenotype <- read_excel(phenotype_path)#%>%filter(!is.na(shell_height))#%>%filter(live_dead_sampling=="L")

cat("Samples with phenotype data:", nrow(phenotype), "\n")
phenotype
```

View distributions. 
```{r}
names(phenotype)
hist(phenotype$condition_index)
hist(phenotype$live_dead_sampling)
hist(phenotype$dry_tissue_weight)
hist(phenotype$cup)
```

# Load resazurin curve features

```{r load-resazurin}
resazurin <- read_csv(resazurin_features_path, show_col_types = FALSE)

cat("Ploidies in resazurin data:    ", n_distinct(resazurin$ploidy), "\n")
```

# Join Resazurin and Field Performance Data

Build a join key from the performance data matching the `trace_id` format in
the curve features (`"YYYYMMDD plate# WELL"`), then merge at the individual
level.

```{r join-data}
# Performance metrics to correlate
perf_metrics <- c("shell_height", "shell_length", "shell_width",
                  "cup", "fan", "whole_wet_weight",
                  "dry_tissue_weight", "condition_index", "live_dead_sampling")

# Resazurin curve feature columns (numeric, excluding bookkeeping)
feature_cols <- resazurin %>%
  select(where(is.numeric)) %>%
  select(-any_of("n_timepoints")) %>%
  names()
feature_cols <- setdiff(feature_cols, "baseline_value")

# Build trace_id in the performance data to match curve_features format
performance_clean <- phenotype %>%
  mutate(
    trace_id = paste(date, plate, well),
    family   = str_to_lower(family),
    ploidy   = str_to_lower(ploidy)
  )

# Inner-join on trace_id — keeps only individuals present in both datasets
joined <- resazurin %>%
  select(trace_id, family_id_group, ploidy, value_metric,
         all_of(feature_cols)) %>%
  inner_join(
    performance_clean %>% select(trace_id, all_of(perf_metrics)),
    by = "trace_id"
  )

cat("Individuals matched:", nrow(joined), "\n")
cat("\nCounts by ploidy x value_metric:\n")
count(joined, ploidy, value_metric) %>% print()

joined<-joined%>%mutate(shell_width=as.numeric(shell_width))

write_csv(joined, file.path(out_dir, "joined_resazurin_performance.csv"))
```

# Diagnose the Join

Check how many individuals matched between datasets and whether the join key
format is consistent.

```{r check-join}
resaz_ids  <- resazurin %>% distinct(trace_id) %>% pull(trace_id)
perf_ids   <- performance_clean %>% distinct(trace_id) %>% pull(trace_id)

cat("Resazurin unique trace_ids:      ", length(resaz_ids), "\n")
cat("Performance unique trace_ids:    ", length(perf_ids),  "\n")
cat("Matched (inner join) rows:       ", nrow(joined),       "\n")
cat("Matched unique individuals:      ", n_distinct(joined$trace_id), "\n")
cat("Match rate (of resazurin IDs):   ",
    round(n_distinct(joined$trace_id) / length(resaz_ids), 2), "\n\n")

cat("Example resazurin trace_ids:\n")
print(head(resaz_ids, 6))

cat("\nExample performance trace_ids:\n")
print(head(perf_ids, 6))

unmatched_resaz <- setdiff(resaz_ids, perf_ids)
unmatched_perf  <- setdiff(perf_ids,  resaz_ids)
cat("\nResazurin IDs absent from performance:", length(unmatched_resaz), "\n")
if (length(unmatched_resaz) > 0) print(head(unmatched_resaz, 10))

cat("\nPerformance IDs absent from resazurin:", length(unmatched_perf), "\n")
if (length(unmatched_perf) > 0) print(head(unmatched_perf, 10))
```

# Reduce Correlated Curve Features

Features that are nearly perfectly collinear carry redundant information and
inflate the multiple-testing burden. A Spearman correlation matrix is
computed across features; hierarchical clustering on 1 − |ρ| groups
redundant features, and the highest-variance representative is kept from each
cluster (cutoff |ρ| > 0.75).

```{r feature-reduction, fig.width=12, fig.height=10}
# Drop features with near-zero variance or too many NAs before correlating
feat_data <- joined %>%
  select(all_of(feature_cols)) %>%
  select(where(~ var(.x, na.rm = TRUE) > 1e-10 && mean(is.finite(.x)) >= 0.5))

# Update feature_cols to only non-degenerate features
feature_cols <- names(feat_data)

# --- correlation matrix ---
feat_mat <- cor(feat_data, use = "pairwise.complete.obs", method = "spearman")

# Replace any remaining NAs in the correlation matrix with 0 (no correlation)
feat_mat[is.na(feat_mat)] <- 0
diag(feat_mat) <- 1

# heatmap of feature autocorrelations
feat_cor_df <- feat_mat %>%
  as.data.frame() %>%
  rownames_to_column("f1") %>%
  pivot_longer(-f1, names_to = "f2", values_to = "r")

p_feat_cor <- ggplot(feat_cor_df, aes(x = f1, y = f2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1), name = "Spearman\nrho") +
  labs(title = "Curve feature autocorrelations",
       x = NULL, y = NULL) +
  theme_classic(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))

p_feat_cor
ggsave(file.path(fig_dir, "feature_autocorrelation.png"),
       p_feat_cor, width = 12, height = 10)

# --- hierarchical clustering to find redundant features ---
dist_feat <- as.dist(1 - abs(feat_mat))
hc_feat   <- hclust(dist_feat, method = "average")
clusters_feat <- cutree(hc_feat, h = 0.25)   # h = 1 - 0.75

feat_vars <- joined %>%
  select(all_of(feature_cols)) %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "variance")

selected_features <- tibble(feature  = names(clusters_feat),
                             cluster  = clusters_feat) %>%
  left_join(feat_vars, by = "feature") %>%
  group_by(cluster) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  pull(feature)

cat("Original features:  ", length(feature_cols),      "\n")
cat("After reduction:    ", length(selected_features),  "\n")
cat("Retained features:\n")
print(sort(selected_features))

# overwrite feature_cols so all downstream chunks use the reduced set
feature_cols <- selected_features
```

# Reduce Correlated Performance Metrics

The same procedure applied to the field performance metrics. Shell dimensions,
weights, and condition index are often strongly co-correlated within a size
cohort; retaining one representative per cluster avoids testing the same
biological signal multiple times.

```{r perf-reduction, fig.width=8, fig.height=7}
# --- correlation matrix ---
perf_data <- joined %>%
  select(all_of(perf_metrics)) %>%
  mutate(across(everything(), as.numeric)) %>%
  select(where(~ var(.x, na.rm = TRUE) > 1e-10 && mean(is.finite(.x)) >= 0.5))

perf_metrics <- names(perf_data)

perf_mat <- cor(perf_data, use = "pairwise.complete.obs", method = "spearman")
perf_mat[is.na(perf_mat)] <- 0
diag(perf_mat) <- 1

perf_cor_df <- perf_mat %>%
  as.data.frame() %>%
  rownames_to_column("m1") %>%
  pivot_longer(-m1, names_to = "m2", values_to = "r")

p_perf_cor <- ggplot(perf_cor_df, aes(x = m1, y = m2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = round(r, 2)), size = 2.5, colour = "grey20") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1), name = "Spearman\nrho") +
  labs(title = "Performance metric autocorrelations",
       x = NULL, y = NULL) +
  theme_classic(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_perf_cor
ggsave(file.path(fig_dir, "perf_metric_autocorrelation.png"),
       p_perf_cor, width = 8, height = 7)

# --- hierarchical clustering ---
dist_perf    <- as.dist(1 - abs(perf_mat))
hc_perf      <- hclust(dist_perf, method = "average")
clusters_perf <- cutree(hc_perf, h = 0.25)

perf_vars <- joined %>%
  select(all_of(perf_metrics)) %>%
  mutate(across(everything(), as.numeric)) %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "variance")

selected_perf_metrics <- tibble(metric   = names(clusters_perf),
                                 cluster  = clusters_perf) %>%
  left_join(perf_vars, by = "metric") %>%
  group_by(cluster) %>%
  slice_max(variance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  pull(metric)

cat("Original performance metrics: ", length(perf_metrics),         "\n")
cat("After reduction:              ", length(selected_perf_metrics), "\n")
cat("Retained metrics:\n")
print(sort(selected_perf_metrics))

# show which cluster each original metric fell into
tibble(metric  = names(clusters_perf),
       cluster = clusters_perf,
       kept    = metric %in% selected_perf_metrics) %>%
  arrange(cluster, metric) %>%
  print()

# overwrite perf_metrics so all downstream chunks use the reduced set
perf_metrics <- selected_perf_metrics
```

# Correlate Resazurin Features with Field Performance

Spearman ρ (rank-based) is the primary statistic; Pearson *r* is reported
alongside. Correlations are computed for every combination of
ploidy × value_metric × resazurin feature × performance metric.
P-values are BH-adjusted within each ploidy × value_metric stratum.

```{r correlations}
cor_one <- function(x, y) {
  ok   <- is.finite(x) & is.finite(y)
  n_ok <- sum(ok)
  if (n_ok < 5 || sd(x[ok]) == 0 || sd(y[ok]) == 0) {
    return(tibble(n = n_ok,
                  spearman_rho = NA_real_, spearman_p = NA_real_,
                  pearson_r    = NA_real_, pearson_p  = NA_real_))
  }
  sp <- suppressWarnings(cor.test(x[ok], y[ok], method = "spearman"))
  pe <- suppressWarnings(cor.test(x[ok], y[ok], method = "pearson"))
  tibble(n            = n_ok,
         spearman_rho = unname(sp$estimate),
         spearman_p   = sp$p.value,
         pearson_r    = unname(pe$estimate),
         pearson_p    = pe$p.value)
}

# Long format: one row per individual x feature x perf_metric combination
joined_long <- joined %>%
  pivot_longer(all_of(feature_cols),
               names_to  = "feature",
               values_to = "feature_value") %>%
  pivot_longer(all_of(perf_metrics),
               names_to  = "perf_metric",
               values_to = "perf_value")

# Correlations — extract vectors inside group_modify to avoid scoping issues
correlations <- joined_long %>%
  filter(is.finite(feature_value), is.finite(perf_value)) %>%
  group_by(ploidy, value_metric, feature, perf_metric) %>%
  group_modify(~ cor_one(.x$feature_value, .x$perf_value)) %>%
  ungroup() %>%
  group_by(ploidy, value_metric) %>%
  mutate(p_adj_bh = p.adjust(spearman_p, method = "BH")) %>%
  ungroup() %>%
  arrange(ploidy, value_metric, desc(abs(spearman_rho)))

write_csv(correlations, file.path(out_dir, "individual_correlations.csv"))

cat("Total correlations computed:", nrow(correlations), "\n")
cat("Significant (BH adj-p < 0.05):",
    sum(correlations$p_adj_bh < 0.05, na.rm = TRUE), "\n\n")

# Top 5 per ploidy x value_metric
correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(ploidy, value_metric) %>%
  slice_max(abs(spearman_rho), n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  select(ploidy, value_metric, feature, perf_metric,
         n, spearman_rho, spearman_p, p_adj_bh) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  print(n = 40)
```

# Leave-One-Individual-Out Cross-Validation

For each top candidate predictor (top 3 by |ρ| within each
ploidy × value_metric × performance-metric stratum), a linear model is fit
on n−1 individuals and the held-out individual is predicted. This
leave-one-individual-out CV (LOIO-CV) tests whether the correlation
between a resazurin feature and field performance generalises beyond the
training individuals. CV Spearman ρ and RMSE summarise predictive accuracy.

```{r loio-fn}
# LOIO-CV for a single feature x perf_metric pair
# df must have columns: x (resazurin feature value), y (performance value)
loio_cv <- function(df) {
  df <- df %>% filter(is.finite(x), is.finite(y))
  n  <- nrow(df)
  if (n < 6 || sd(df$x) == 0) {
    return(tibble(n = n, cv_spearman = NA_real_, cv_rmse = NA_real_))
  }
  preds <- map_dbl(seq_len(n), function(i) {
    fit <- lm(y ~ x, data = df[-i, ])                   # train on n-1
    predict(fit, newdata = df[i, , drop = FALSE])        # predict held-out
  })
  cv_sp <- suppressWarnings(cor(preds, df$y, method = "spearman"))
  tibble(n           = n,
         cv_spearman = cv_sp,
         cv_rmse     = sqrt(mean((preds - df$y)^2)))
}
```

```{r loio-candidates}
# Top 3 features per ploidy x value_metric x perf_metric as CV candidates
candidates <- correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(ploidy, value_metric, perf_metric) %>%
  slice_max(abs(spearman_rho), n = 3, with_ties = FALSE) %>%
  ungroup() %>%
  select(ploidy, value_metric, feature, perf_metric, spearman_rho, p_adj_bh)

# Run LOIO-CV for every candidate
loio_results <- joined_long %>%
  inner_join(candidates,
             by = c("ploidy", "value_metric", "feature", "perf_metric")) %>%
  filter(is.finite(feature_value), is.finite(perf_value)) %>%
  rename(x = feature_value, y = perf_value) %>%
  group_by(ploidy, value_metric, feature, perf_metric, spearman_rho, p_adj_bh) %>%
  group_modify(~ loio_cv(.x)) %>%
  ungroup() %>%
  arrange(ploidy, value_metric, perf_metric, desc(abs(cv_spearman)))

write_csv(loio_results, file.path(out_dir, "loio_cv_performance.csv"))

cat("LOIO-CV results\n")
loio_results %>%
  select(ploidy, value_metric, feature, perf_metric,
         n, spearman_rho, cv_spearman, cv_rmse) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  print(n = 60)
```

```{r loio-plot, fig.width=14, fig.height=10}
p_loio <- loio_results %>%
  filter(is.finite(cv_spearman)) %>%
  mutate(
    feature = reorder(feature, abs(cv_spearman)),
    sig     = p_adj_bh < 0.05
  ) %>%
  ggplot(aes(x = cv_spearman, y = feature, fill = sig)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  facet_grid(perf_metric ~ ploidy + value_metric, scales = "free_y") +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#0072B2"),
    name   = "Correlation BH-adj p < 0.05"
  ) +
  labs(
    x     = "CV Spearman rho (LOIO)",
    y     = "Resazurin feature",
    title = "Leave-one-individual-out CV: predictive performance by ploidy"
  ) +
  theme_classic(base_size = 9) +
  theme(strip.text = element_text(size = 7))

p_loio
ggsave(file.path(fig_dir, "loio_cv_performance.png"),
       p_loio, width = 14, height = 10)
```

# Correlation Heatmap

Heatmap of Spearman ρ for every resazurin feature × performance metric pair,
faceted by ploidy and value metric. Significance stars use BH-adjusted p-values.

```{r cor-heatmap, fig.width=14, fig.height=12}
heatmap_df <- correlations %>%
  filter(is.finite(spearman_rho)) %>%
  mutate(
    sig_label = case_when(
      spearman_p < 0.001 ~ "***",
      spearman_p < 0.01  ~ "**",
      spearman_p < 0.05  ~ "*",
      TRUE             ~ ""
    )
  )

p_heatmap <- ggplot(heatmap_df,
    aes(x = perf_metric, y = feature, fill = spearman_rho)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = sig_label), size = 2.8, colour = "black") +
  facet_grid(value_metric ~ ploidy) +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#B2182B",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Spearman\nrho"
  ) +
  labs(
    x       = "Field performance metric",
    y       = "Resazurin curve feature",
    title   = "Resazurin feature vs field performance correlations by ploidy",
    caption = "* Spearman p < 0.05  ** < 0.01  *** < 0.001"
  ) +
  theme_classic(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7),
    strip.text  = element_text(size = 9)
  )

p_heatmap
ggsave(file.path(fig_dir, "resazurin_performance_heatmap.png"),
       p_heatmap, width = 14, height = 12)
```

# Correlation Ranking Plot

Top 8 resazurin features by |ρ| for each performance metric, split by ploidy.

```{r cor-ranking-plot, fig.width=14, fig.height=10}
rank_df <- correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(ploidy, value_metric, perf_metric) %>%
  slice_max(abs(spearman_rho), n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    feature = reorder(feature, abs(spearman_rho)),
    sig     = spearman_p < 0.05
  )

p_rank <- ggplot(rank_df,
    aes(x = spearman_rho, y = feature, fill = sig)) +
  geom_col(width = 0.75) +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  facet_grid(perf_metric ~ ploidy + value_metric, scales = "free_y") +
  scale_fill_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#0072B2"),
    name   = "Spearman p < 0.05"
  ) +
  labs(
    x     = "Spearman rho",
    y     = "Resazurin curve feature",
    title = "Top resazurin features correlated with each field performance metric"
  ) +
  theme_classic(base_size = 9) +
  theme(strip.text = element_text(size = 7))

p_rank
ggsave(file.path(fig_dir, "resazurin_performance_ranking.png"),
       p_rank, width = 14, height = 10)
```

# Scatter Plots: Top Correlations per Ploidy

Top 4 feature × performance metric pairs per ploidy × value_metric, with
individual data points and linear trend.

```{r top-scatter, fig.width=14, fig.height=10}
top_pairs <- correlations %>%
  filter(is.finite(spearman_rho)) %>%
  group_by(ploidy, value_metric) %>%
  slice_max(abs(spearman_rho), n = 4, with_ties = FALSE) %>%
  ungroup()

scatter_df <- joined_long %>%
  inner_join(
    top_pairs %>% select(ploidy, value_metric, feature, perf_metric,
                         spearman_rho, spearman_p),
    by = c("ploidy", "value_metric", "feature", "perf_metric")
  ) %>%
  filter(is.finite(feature_value), is.finite(perf_value)) %>%
  mutate(
    panel_label = sprintf("%s ~ %s\n%s | %s  rho=%.2f",
                          feature, perf_metric, ploidy,
                          value_metric, spearman_rho)
  )

p_scatter <- ggplot(scatter_df,
    aes(x = feature_value, y = perf_value, colour = ploidy)) +
  geom_point(alpha = 0.6, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 0.8) +
  facet_wrap(~ panel_label, scales = "free", ncol = 4) +
  scale_colour_manual(
    values = c("diploid"  = "#0072B2",
               "triploid" = "#D55E00"),
    name   = "Ploidy"
  ) +
  labs(
    x     = "Resazurin feature value",
    y     = "Field performance metric value",
    title = "Top resazurin–performance correlations (individual level)"
  ) +
  theme_classic(base_size = 9) +
  theme(strip.text = element_text(size = 7))

p_scatter
ggsave(file.path(fig_dir, "top_correlation_scatter.png"),
       p_scatter, width = 14, height = 10)
```

# Diploid vs Triploid Comparison

Do the same resazurin features predict field performance in both ploidies, or
does predictive power diverge? Points off the diagonal predict one ploidy much
better than the other.

```{r ploidy-comparison-plot, fig.width=10, fig.height=8}
ploidy_wide <- correlations %>%
  filter(is.finite(spearman_rho)) %>%
  select(ploidy, value_metric, feature, perf_metric, spearman_rho) %>%
  pivot_wider(names_from = ploidy, values_from = spearman_rho) %>%
  filter(is.finite(diploid), is.finite(triploid)) %>%
  mutate(
    rho_diff    = triploid - diploid,
    abs_max     = pmax(abs(diploid), abs(triploid)),
    label_feat  = if_else(abs_max > 0.3 | abs(rho_diff) > 0.2,
                          paste(feature, perf_metric, sep = "\n"),
                          NA_character_)
  )

p_ploidy_compare <- ggplot(ploidy_wide,
    aes(x = diploid, y = triploid, colour = perf_metric)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.25, colour = "grey70") +
  geom_vline(xintercept = 0, linewidth = 0.25, colour = "grey70") +
  geom_point(size = 2.2, alpha = 0.75) +
  ggrepel::geom_text_repel(aes(label = label_feat),
                            size = 2.8, max.overlaps = 15,
                            show.legend = FALSE, na.rm = TRUE) +
  facet_wrap(~ value_metric) +
  labs(
    x        = "Spearman rho (diploid)",
    y        = "Spearman rho (triploid)",
    colour   = "Performance metric",
    title    = "Resazurin predictive power: diploid vs triploid",
    subtitle = "Points on the diagonal predict both ploidies equally"
  ) +
  theme_classic(base_size = 11)

p_ploidy_compare
ggsave(file.path(fig_dir, "diploid_vs_triploid_correlation_comparison.png"),
       p_ploidy_compare, width = 10, height = 8)
```

# Significant Correlations Summary

```{r sig-summary, results='asis'}
sig_table <- correlations %>%
  filter(spearman_p < 0.05, is.finite(spearman_rho)) %>%
  arrange(ploidy, desc(abs(spearman_rho))) %>%
  transmute(
    ploidy, value_metric, feature, perf_metric, n,
    spearman_rho = round(spearman_rho, 3),
    spearman_p   = signif(spearman_p, 3),
    p_adj_bh     = signif(p_adj_bh, 3)
  )

cat("## Significant correlations (BH-adjusted p < 0.05)\n\n")

if (nrow(sig_table) == 0) {
  cat("No significant correlations found at this threshold.\n\n")
} else {
  cat(knitr::kable(sig_table, format = "pipe"), sep = "\n")
}
```

# Session info

```{r session-info}
sessionInfo()
```
